---
title: "Census Data in R"
output: html_notebook
---


Focused on the census data from `tidycensus`. 
API key required via census.gov.

```{r setup}
library(tidyverse)  # Oppan Tidy Style
library(tidycensus) # Census.gov API
library(tigris)     # TIGRIS shapefiles
```

# Census data in R with tidycensus

`get_decennial` requires the geography (level of aggregation) and variables 
parameters.

`get_acs` - American Community Survey data - based on samples, or 
estimates of the population. Dataframes returned from 
this function will have estimates and moe (margin of error for 90% CI).

```{r, eval=FALSE, include=FALSE}
Sys.getenv("CENSUS_API_KEY")
```

```{r}
# Obtain and view state populations from the 2010 US Census
state_pop <- get_decennial(geography = "state", variables = "P001001")
state_pop
```

```{r}
# Obtain and view state median household income from the 2012-2016 American Community Survey
state_income <- get_acs(geography = "state", variables = "B19013_001")

state_income
```

Geography regions
- legal entities (e.g. country)
- statistical etities (e.g. tract)

`tidycensus` docs include geography available
`tidycensus` "returns data in tidy foramt, but can return data in wide format with the 
    `output = "wide"` parameter.
    
```{r}
# Get an ACS dataset for Census tracts in Texas by setting the state
# this is the median household income
tx_income <- get_acs(geography = "tract",
                     variables = c(hhincome = "B19013_001"),
                     state = "TX")

# Inspect the dataset
head(tx_income)

# Get an ACS dataset for Census tracts in Travis County, TX
travis_income <- get_acs(geography = "tract",
                         variables = "B19013_001", 
                         state = "WA",
                         county = "King")

# Inspect the dataset
head(travis_income)
```
    

```{r}
# Return county data in wide format
or_wide <- get_acs(geography = "county", 
                     state = "OR",
                     variables = c(hhincome = "B19013_001", 
                            medage = "B01002_001"), 
                     output = "wide")

# Compare output to the tidy format from previous exercises
head(or_wide)

# Create a scatterplot
plot(or_wide$hhincomeE, or_wide$medageE)
```

To find census variables, look at online resources like Census Reporter - https://censusreporter.org/A

`load_variables` helps to find variables. takes year, dataset (`acs5`) and 
cache boolean, this can then be explored. Returns name, label, and concept 
fields.

ACS Variable structure
B19001_002e
B is a base (most data), (C collapsed, DP data profile, or S subject)
19001 is the table ID
002 the variable code within the table
E is the estimate (optional in tidycensus, returns both the *E*stimate and *M*argin of error)

```{r}
# Load variables from the 2012-2016 ACS
v16 <- load_variables(year = 2016,
           dataset = "acs5",
           cache = TRUE)
# Get variables from the ACS Data Profile
v16p <- load_variables(year = 2016,
                       dataset = "acs5/profile",
                       cache = TRUE)
# Set year and dataset to get variables from the 2000 Census SF3
v00 <- load_variables(year = 2000,
                      dataset = "sf3",
                      cache = TRUE)
```






```{r}
# Filter for table B19001
filter(v16, str_detect(name, "B19001"))

# Use public transportation to search for related variables
filter(v16p, str_detect(label, fixed("public transportation", 
                                ignore_case = TRUE)))
```
```{r}
get_acs(geography = "county", state = "WA", year = 2016, 
        variables = "DP03_0021P", survey = "acs1") %>% 
  mutate(NAME = gsub(" County, Washington", "", NAME)) %>% 
  ggplot(aes(estimate, reorder(NAME, estimate))) + 
  geom_point(size = 3, color = "navy") + 
  geom_errorbarh(aes(xmax = estimate + moe, xmin = estimate - moe)) + 
  scale_x_continuous(labels = scales::percent_format(scale = 1, accuracy = 1)) +
  hrbrthemes::theme_ipsum() + 
  labs(title = "Percent of workers commuting via public transit", 
       caption = "2016 ACS",
       x = "Percent",
       y = "County")

```

# Tables and summary variables in tidycensus

`get_acs` and `get_decennial` have an optional `table` param that allows all the 
variables to be fetched at once without specifying a variable param.

```{r}

# Download table "B19001"
wa_income <- get_acs(geography = "county", 
                 state = "WA", 
                 table = "B19001")

# Check out the first few rows of wa_income
wa_income
```

```{r}
# Assign Census variables vector to race_vars  
race_vars <- c(White = "B03002_003", Black = "B03002_004", Native = "B03002_005", 
               Asian = "B03002_006", HIPI = "B03002_007", Hispanic = "B03002_012")

# Request a summary variable from the ACS
ca_race <- get_acs(geography = "county", 
                   state = "WA",
                   variables = race_vars, 
                   summary_var = "B03002_001")

# Calculate a new percentage column and check the result
ca_race_pct <- ca_race %>%
  mutate(pct = 100 * (estimate / summary_est))

ca_race_pct
```


```{r}
# Group the dataset and filter the estimate
ca_largest <- ca_race %>%
  group_by(GEOID) %>%
  filter(estimate == max(estimate)) 

head(ca_largest)

# Group the dataset and get a breakdown of the results
ca_largest %>% 
  group_by(variable) %>% 
  tally()
```

```{r}
# Use a tidy workflow to wrangle ACS data
wa_grouped <- wa_income %>%
  filter(variable != "B19001_001") %>%
  mutate(incgroup = case_when(
    variable < "B19001_008" ~ "below35k", 
    variable < "B19001_013" ~ "35kto75k", 
    TRUE ~ "above75k"
  )) %>%
  group_by(NAME, incgroup) %>%
  summarize(group_est = sum(estimate))

wa_grouped
```

```{r}
# Map through ACS1 estimates to see how they change through the years
mi_cities <- map_df(2012:2016, function(x) {
  get_acs(geography = "place", 
          variables = c(totalpop = "B01003_001"), 
          state = "MI", 
          survey = "acs1", 
          year = x) %>%
    mutate(year = x)
})

mi_cities %>% arrange(NAME, year)
```
```{r}
# Get data on elderly poverty by Census tract in Vermont
vt_eldpov <- get_acs(geography = "tract", 
                     variables = c(eldpovm = "B17001_016", 
                                   eldpovf = "B17001_030"), 
                     state = "VT")

vt_eldpov

# Identify rows with greater margins of error than their estimates
moe_check <- filter(vt_eldpov, moe > estimate)

# Check proportion of rows where the margin of error exceeds the estimate
nrow(moe_check) / nrow(vt_eldpov)
```

```{r}
# Group the dataset and calculate a derived margin of error
vt_eldpov2 <- vt_eldpov %>%
  group_by(GEOID) %>%
  summarize(
    estmf = sum(estimate), 
    moemf = moe_sum(moe= moe, estimate = estimate)
  )

# Filter rows where newly-derived margin of error exceeds newly-derived estimate
moe_check2 <- filter(vt_eldpov2, moemf > estmf)

# Check proportion of rows where margin of error exceeds estimate
nrow(moe_check2) / nrow(vt_eldpov2)
```
```{r}
# Request median household income data
maine_inc <- get_acs(geography = "county", 
                     variables = c(hhincome = "B19013_001"), 
                     state = "ME") 

# Generate horizontal error bars with dots
ggplot(maine_inc, aes(x = estimate, y = NAME)) + 
  geom_errorbarh(aes(xmin = estimate - moe, xmax = estimate + moe)) + 
  geom_point()
```

```{r}
# Remove unnecessary content from the county's name
maine_inc2 <- maine_inc %>%
  mutate(NAME = str_replace(NAME, " County, Maine", ""))

# Build a margin of error plot incorporating your modifications
ggplot(maine_inc2, aes(x = estimate, y = reorder(NAME, estimate))) + 
  geom_errorbarh(aes(xmin = estimate - moe, xmax = estimate + moe)) + 
  geom_point(size = 3, color = "darkgreen") + 
  theme_grey(base_size = 14) + 
  labs(title = "Median household income", 
       subtitle = "Counties in Maine", 
       x = "ACS estimate (bars represent margins of error)", 
       y = "") + 
  scale_x_continuous(labels = scales::dollar)
```


# Understanding Census geography and tigris basics

TIGER - Topologically integrated geologically encodind databse

`tigris` downloads line and shapes from Census Beuro and pulls them into R as Spatial objects.

Include geographic and tract regions.
Also includes features such as roads and water features.
```{r}

# Get a counties dataset for Colorado and plot it
co_counties <- counties(state = "CO")
head(co_counties)
```
```{r}
# Get a Census tracts dataset for Denver County, Colorado and plot it
denver_tracts <- tracts(state = "CO", county = "Denver")
plot(denver_tracts)
king_tracts <- tracts(state="WA", county = "King")
plot(king_tracts)
```


```{r}
# Plot area water features for Lane County, Oregon
lane_water <- area_water(state = "OR", county = "Lane")
plot(lane_water)
king_water <- area_water(state = "WA", county = "King")
plot(king_water)
```

```{r}
# Plot primary & secondary roads for the state of New Hampshire
nh_roads <- primary_secondary_roads(state = "NH")
plot(nh_roads)
wa_roads <- primary_secondary_roads(state = "WA")
plot(wa_roads)
```

Looking at structure of the Spatial (sp) objects

```{r}
# Check the class of the data
class(co_counties)

# Take a look at the information in the data slot
head(co_counties@data)

# Check the coordinate system of the data
co_counties@proj4string
```

Difference between cartogragphic and tigris boundaries

```{r}
# Get a counties dataset for Michigan
mi_tiger <- counties("MI")

# Get the equivalent cartographic boundary shapefile
mi_cb <- counties("MI", cb = TRUE)

# Overlay the two on a plot to make a comparison
plot(mi_tiger$geometry)
plot(mi_cb$geometry, add = TRUE, border = "red")
```

```{r}
# Get a historic Census tract shapefile from 1990 for Williamson County, Texas
williamson90 <- tracts(state = "TX", county = "Williamson", 
                       cb = TRUE, year = 1990)

# Compare with a current dataset for 2016
williamson16 <- tracts(state = "TX", county = "Williamson", 
                       cb = TRUE, year = "2016")

# Plot the geometry to compare the results                       
par(mfrow = c(1, 2))
plot(williamson90$geometry)
plot(williamson16$geometry)
```

```{r}
# Get Census tract boundaries for Oregon and Washington
or_tracts <- tracts("OR", cb = TRUE)
wa_tracts <- tracts("WA", cb = TRUE)

# Check the tigris attributes of each object
attr(or_tracts, "tigris")
attr(wa_tracts, "tigris")

# Combine the datasets then plot the result
or_wa_tracts <- rbind_tigris(or_tracts, wa_tracts)
plot(or_wa_tracts$geometry)
```

Tidy patterns can also be applied to combine multiple tigris (sf) objects 
together.

```{r}
# Generate a vector of state codes and assign to new_england
new_england <- c("ME", "NH", "VT", "MA")

# Iterate through the states and request tract data for state
ne_tracts <- map(new_england, function(x) {
  tracts(state = x, cb = TRUE)
}) %>%
  rbind_tigris()

plot(ne_tracts$geometry)
```
```{r}
tracts(state = "WA", house = "upper")
```


```{r}
# Get boundaries for Texas and set the house parameter
tx_house <- state_legislative_districts(state = "TX", house = "lower", cb = TRUE)

# Merge data on legislators to their corresponding boundaries
tx_joined <- left_join(tx_house, tx_members, by = c("NAME" = "District"))

head(tx_joined)
```

```{r}
state_legislative_districts(state = "WA", cb = TRUE) -> wa_senate
plot(wa_senate$geometry)
ggplot(wa_senate) + geom_sf() + 
  coord_sf(datum = NA) + 
  theme_minimal()
```

```{r}
# Plot the legislative district boundaries
ggplot(tx_joined) + 
  geom_sf()

# Set fill aesthetic to map areas represented by Republicans and Democrats
ggplot(tx_joined, aes(fill = Party)) + 
  geom_sf()

# Set values so that Republican areas are red and Democratic areas are blue
ggplot(tx_joined, aes(fill = Party)) + 
  geom_sf() + 
  scale_fill_manual(values = c("R" = "red", "D" = "blue"))
```

```{r}
# Draw a ggplot without gridlines and with an informative title
ggplot(tx_joined, aes(fill = Party)) + 
  geom_sf() + 
  coord_sf(crs = 3083, datum = NA) + 
  scale_fill_manual(values = c("R" = "red", "D" = "blue")) + 
  theme_minimal(base_size = 16) + 
  labs(title = "State House Districts in Texas")
```

```{r}
library(sf)

# Get dataset with geometry set to TRUE
orange_value <- get_acs(geography = "tract", state = "CA", 
                    county = "Orange", 
                    variables = "B25077_001", 
                    geometry = TRUE)

# Plot the estimate to view a map of the data
plot(orange_value["estimate"])
```

```{r}
# Get an income dataset for Idaho by school district
idaho_income <- get_acs(geography = "school district (unified)", 
                        variables = "B19013_001", 
                        state = "ID")

# Get a school district dataset for Idaho
idaho_school <- school_districts(state = "ID", type = "unified", class = "sf")

# Join the income dataset to the boundaries dataset
id_school_joined <- left_join(idaho_school, idaho_income, idaho_school, by = "GEOID")

plot(id_school_joined["estimate"])
```
```{r}
# Get a dataset of median home values from the 1-year ACS
state_value <- get_acs(geography = "state", 
                       variables = "B25077_001", 
                       survey = "acs1", 
                       geometry = TRUE, 
                       shift_geo = TRUE)

# Plot the dataset to view the shifted geometry
plot(state_value["estimate"])
```

```{r}
# Set the color guide to FALSE and add a subtitle and caption to your map
ggplot(marin_value, aes(fill = estimate, color = estimate)) + 
  geom_sf() + 
  scale_fill_viridis_c(labels = scales::dollar) +  
  scale_color_viridis_c(guide = FALSE) + 
  theme_minimal() + 
  coord_sf(crs = 26911, datum = NA) + 
  labs(title = "Median owner-occupied housing value by Census tract", 
       subtitle = "Marin County, California", 
       caption = "Data source: 2012-2016 ACS.\nData acquired with the R tidycensus package.", 
       fill = "ACS estimate")
```

```{r}
# Generate point centers
centers <- st_centroid(state_value)

# Set size parameter and the size range
ggplot() + 
  geom_sf(data = state_value, fill = "white") + 
  geom_sf(data = centers, aes(size = estimate), shape = 21, 
          fill = "lightblue", alpha = 0.7, show.legend = "point") + 
  scale_size_continuous(range = c(1, 20))
```

Other pacakges 
- censusapi  - full census API access, very extensive
- ipumsr - historical data
- cancensus - Canadian census

Other Recommendations
- Working with data in the tidyverse
- data visualizeation with ggplot2
- interactive maps with leaflet in R
