---
title: "Machine Learning in the Tidyverse"
output: html_notebook
---

```{r setup}
library(tidyverse)  # Oppan Tidy Style
library(gapminder)  # Primary data set for examples
library(rsample)    # Successor to caret - tidy style
library(Metrics)    # Dima's favored metric is MAE, mean absolute error, contained in this package
```

```{r}
# Explore gapminder
head(gapminder)

# Prepare the nested dataframe gap_nested
library(tidyverse)
gap_nested <- gapminder %>% 
  group_by(country) %>% 
  nest()

# Explore gap_nested
head(gap_nested)
```

## training, testing and validating

Uses `rsample`. Yay!
```{r}
set.seed(42)

# Prepare the initial split object
gap_split <- initial_split(gapminder, prop = 0.75)

# Extract the training dataframe
training_data <- training(gap_split)

# Extract the testing dataframe
testing_data <- testing(gap_split)

# Calculate the dimensions of both training_data and testing_data
dim(training_data)
dim(testing_data)
```

Cross-validation with R sample. `vfold_cv` does repeated splits of the 
data frame, from which training and validation (testing) subframes can be 
extracted using the `training` and `testing` functions.

```{r}
set.seed(42)

# Prepare the dataframe containing the cross validation partitions
cv_split <- vfold_cv(training_data, v = 5)

cv_data <- cv_split %>% 
  mutate(
    # Extract the train dataframe for each split
    train = map(splits, ~training(.x)), 
    # Extract the validate dataframe for each split
    validate = map(splits, ~testing(.x))
  )

# Use head() to preview cv_data
head(cv_data)
```
```{r}
# Build a model using the train data for each fold of the cross validation
cv_models_lm <- cv_data %>% 
  mutate(model = map(train, ~lm(formula = lifeExp~., data = .x)))
```
```{r}
cv_prep_lm <- cv_models_lm %>% 
  mutate(
    # Extract the recorded life expectancy for the records in the validate dataframes
    validate_actual = map(validate, ~.x$lifeExp),
    # Predict life expectancy for each validate set using its corresponding model
    validate_predicted = map2(.x = model, .y = validate, ~predict(.x, .y))
  )
```

```{r}
# Calculate the mean absolute error for each validate fold       
cv_eval_lm <- cv_prep_lm %>% 
  mutate(validate_mae = map2_dbl(validate_actual,validate_predicted, ~mae(actual = .x, predicted = .y)))

# Print the validate_mae column
cv_eval_lm$validate_mae

# Calculate the mean of validate_mae column
mean(cv_eval_lm$validate_mae)
```

```{r}
library(ranger)

# Build a random forest model for each fold
cv_models_rf <- cv_data %>% 
  mutate(model = map(train, ~ranger(formula = lifeExp~., data = .x,
                                    num.trees = 100, seed = 42)))

# Generate predictions using the random forest model
cv_prep_rf <- cv_models_rf %>% 
  mutate(validate_predicted = map2(.x = model, .y = validate, ~predict(.x, .y)$predictions))
```

```{r}
library(ranger)

# need to add validate_actual
cv_prep_rf <- cv_prep_rf %>% mutate(validate_actual = map(validate, ~ .x$lifeExp))

# Calculate validate MAE for each fold
cv_eval_rf <- cv_prep_rf %>% 
  mutate(validate_mae = map2_dbl(validate_actual, validate_predicted, ~mae(actual = .x, predicted = .y)))

# Print the validate_mae column
cv_eval_rf$validate_mae

# Calculate the mean of validate_mae column
mean(cv_eval_rf$validate_mae)
```
```{r}
# Prepare for tuning your cross validation folds by varying mtry
cv_tune <- cv_data %>% 
  crossing(mtry = 2:5) 


# Build a model for each fold & mtry combination
cv_model_tunerf <- cv_tune %>% 
  mutate(model = map2(.x = train, .y = mtry, ~ranger(formula = lifeExp~., 
                                           data = .x, mtry = .y, 
                                           num.trees = 100, seed = 42)))
```

```{r}
# Generate validate predictions for each model
cv_prep_tunerf <- cv_model_tunerf %>% 
  mutate(validate_predicted = map2(.x = model, .y = validate, ~predict(.x, .y)$predictions))

# need to add validate_actual
cv_prep_tunerf <- cv_prep_tunerf %>% mutate(validate_actual = map(validate, ~ .x$lifeExp))

# Calculate validate MAE for each fold and mtry combination
cv_eval_tunerf <- cv_prep_tunerf %>% 
  mutate(validate_mae = map2_dbl(.x = validate_actual, .y = validate_predicted, ~mae(actual = .x, predicted = .y)))

# Calculate the mean validate_mae for each mtry used  
cv_eval_tunerf %>% 
  group_by(mtry) %>% 
  summarise(mean_mae = mean(validate_mae))
```

```{r}
# Build the model using all training data and the best performing parameter
best_model <- ranger(formula = lifeExp~., data = training_data,
                     mtry = 4, num.trees = 100, seed = 42)

# Prepare the test_actual vector
test_actual <- testing_data$lifeExp

# Predict life_expectancy for the testing_data
test_predicted <- predict(best_model, testing_data)$predictions

# Calculate the test MAE
mae(test_actual, test_predicted)
```

## Build, Tune & Evaluate Classification Models

```{r}
set.seed(42)

# Prepare the initial split object
data_split <- initial_split(attrition, prop = 0.75)

# Extract the training dataframe
training_data <- training(data_split)

# Extract the testing dataframe
testing_data <- testing(data_split)
```

```{r}
set.seed(42)
cv_split <- vfold_cv(training_data, v = 5)

cv_data <- cv_split %>% 
  mutate(
    # Extract the train dataframe for each split
    train = map(splits, ~training(.x)),
    # Extract the validate dataframe for each split
    validate = map(splits, ~testing(.x))
  )
```
```{r}
# Build a model using the train data for each fold of the cross validation
cv_models_lr <- cv_data %>% 
  mutate(model = map(train, ~glm(formula = Attrition~., 
                               data = .x, family = "binomial")))
```

Accuracy - How often the model is at predicting true positives and true negatives
Precision - How often is the model correct at the TRUE class (How safe is a true prediction?)
Recall - How often is the model capturing the true class given the number of actual frequency of corrects (How comprehensive are the true predictions?)

```{r}
# Extract the first model and validate 
model <- cv_models_lr$model[[1]]
validate <- cv_models_lr$validate[[1]]

# Prepare binary vector of actual Attrition values in validate
validate_actual <- validate$Attrition == "Yes"

# Predict the probabilities for the observations in validate
validate_prob <- predict(model, validate, type = "response")

# Prepare binary vector of predicted Attrition values for validate
validate_predicted <- validate_prob > 0.5
```

```{r}
library(Metrics)

# Compare the actual & predicted performance visually using a table
table(validate_actual, validate_predicted)

# Calculate the accuracy
accuracy(validate_actual, validate_predicted)

# Calculate the precision
precision(validate_actual, validate_predicted)

# Calculate the recall
recall(validate_actual, validate_predicted)
```
```{r}
cv_prep_lr <- cv_models_lr %>% 
  mutate(
    # Prepare binary vector of actual Attrition values in validate
    validate_actual = map(validate, ~.x$Attrition == "Yes"),
    # Prepare binary vector of predicted Attrition values for validate
    validate_predicted = map2(.x = model, .y = validate, ~predict(.x, .y, type = "response") > 0.5)
  )
```

```{r}
# Calculate the validate recall for each cross validation fold
cv_perf_recall <- cv_prep_lr %>% 
  mutate(validate_recall = map2_dbl(validate_actual, validate_predicted, 
                                    ~recall(actual = .x, predicted = .y)))

# Print the validate_recall column
cv_perf_recall$validate_recall

# Calculate the average of the validate_recall column
mean(cv_perf_recall$validate_recall)
```
```{r}
library(ranger)

# Prepare for tuning your cross validation folds by varying mtry
cv_tune <- cv_data %>%
  crossing(mtry = c(2, 4, 8, 16)) 

# Build a cross validation model for each fold & mtry combination
cv_models_rf <- cv_tune %>% 
  mutate(model = map2(train, mtry, ~ranger(formula = Attrition~., 
                                           data = .x, mtry = .y,
                                           num.trees = 100, seed = 42)))
```

```{r}
cv_prep_rf <- cv_models_rf %>% 
  mutate(
    # Prepare binary vector of actual Attrition values in validate
    validate_actual = map(validate, ~.x$Attrition == "Yes"),
    # Prepare binary vector of predicted Attrition values for validate
    validate_predicted = map2(.x = model, .y = validate, ~predict(.x, .y, type = "response")$predictions == "Yes")
  )

# Calculate the validate recall for each cross validation fold
cv_perf_recall <- cv_prep_rf %>% 
  mutate(recall = map2_dbl(.x = validate_actual, .y = validate_predicted, ~recall(actual = .x, predicted = .y)))

# Calculate the mean recall for each mtry used  
cv_perf_recall %>% 
  group_by(mtry) %>% 
  summarise(mean_recall = mean(recall))
```

```{r}
# Build the logistic regression model using all training data
best_model <- glm(formula = Attrition~., 
                  data = training_data, family = "binomial")


# Prepare binary vector of actual Attrition values for testing_data
test_actual <- testing_data$Attrition == "Yes"

# Prepare binary vector of predicted Attrition values for testing_data
test_predicted <- predict(best_model, testing_data, type = "response") > 0.5
```

```{r}
# Compare the actual & predicted performance visually using a table
table(test_actual, test_predicted)

# Calculate the test accuracy
accuracy(test_actual, test_predicted)

# Calculate the test precision
precision(test_actual, test_predicted)

# Calculate the test recall
recall(test_actual, test_predicted)
```

